\LARGE \bf Bringing robotics into formal education \\ using the Thymio open source hardware robot

### 2016-10-20 paper_ram.tex
final changes

\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools,'' and by the EU-FP7 project ASSISIbf, no. 601074 (survey on open hardware).
Many thanks to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland.
        {General rule for EPFL emails: \tt\small firstname.lastname@epfl.ch}}%
\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland.
        {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
}

@@

Many publications present educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}. 
Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
As a result, 90\% of publications about validation of educational results have been based on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ US\$400) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment. 
%NOTE: SM: commented WeDo to save space and streamline story
%At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
%The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
%In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
%However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
Among the other players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49) and compatible with LEGO\textsuperscript{\textregistered} bricks. 
The low price has pushed drastic design choices: small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
These choices strongly limit its possible usage.
%Among widely used robots, none is part of an open source hardware project.

@@

Teachers, from primary school up to high school, are a primary target user group of the Thymio project.
They decide which tools are used in their class and are key people in the education ecosystem.
For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
Among them, the availability of materials and training plays a key role.
%The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
%Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
%These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
%However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
The development of educational material and courses to train teachers requires a huge effort, based on a good mix between robotics and educational skills. 
Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
%For example, in Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
%Educational material and the related training must therefore be adapted to each situation.
Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
A crowd-sourcing approach may solve this problem; An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO\textsuperscript{\textregistered} itself is moving this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction similar to that of open hardware projects.
An open source community regrouping developers, manufacturers, and end users is therefore a very interesting model to address the distributed development and sharing of educational material and the diffusion of training sessions.
%This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
In this paper, we study a case of implementation of this model.

@@

The resulting Thymio robot possesses a large number of simple sensors:
seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
%A microphone can record sounds and detect claps. 
%Finally, a thermistor measures temperature.
Most of these devices cost less then US\$0.20, the most expensive being the accelerometer with a cost of about US\$0.80, which is an acceptable price given the possibilities it brings to the robot.
%The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need for additional encoders.

@@

Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
Full assembly is required in order for the robots to be certified for use by children. 
%A final assembly by the user is excluded by the certification required in the case of use by children. 
As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
We therefore decided to produce in China, where manual work is less expensive.
%This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
%Because manual work is expensive in western countries, we produce through subcontractors in China.
%Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
%We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
%The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
%Moreover, when selecting features we took into consideration the complexity of the injection mold.
%The design of the mechanical parts has been aligned with these choicesthat are injected, resulting in all custom parts of Thymio costing less than 4\,\$.
%In these conditions Thymio costs 35\,\$. 
The strict quality control, the production management, the support to the users, and the margins for distributors result in a final selling price of US\$130.

@@

Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne.\footnote{\url{http://www.ecal.ch}}
The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
These latter two robots implicitly target a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience.
%The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.

@@

\subsection{Programming Environment}
\label{sec:aseba}

@@

\section{Open source hardware, choices and impact}

The last but not least design choice among the seven mentioned in the previous section is the open source hardware design.
This choice has an impact on the robot design and the way this device is used in the community of users. 
In this section, we analyze in more detail the implications of this choice in the context of educational robotics.
We compare these implications with the results of a survey that returned 35 answers from 11 project leaders, 13 core design team members, 8 contributors, and 3 enthusiastic users of open hardware projects.
%people active in various open source hardware projects based on a worldwide call for contributions. 
%Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
%Although 35 answer are not many, the high proportion of project leaders and core design members allows to extract some meaningful observations.
%54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
%For instance, it is interesting to observe that 34 of the 35 answers come from male respondents.

@@

%We can distinguish two levels of motivation, the institutional and the personal one.
%The institution initiating a project is mainly looking for recognition and/or money. 
Our group has good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware strategy.
Both targeted similar users and have been sold in similar quantities.
What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was associated with the name of the company producing it, not the university that developed it (Fig.~\ref{fig:khepuck}).
The e-puck robot, with open hardware and an image better linked to the university, generated no income for the university but much more visibility.
%Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
In the case of Thymio, the institutional motivation was toward visibility more than money. 
Therefore, from an institutional point of view, the open source hardware strategy seemed more adapted to the desired outcomes.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/others_robots}
\caption{Comparison between e-Puck and Khepera in terms of mentions in Google Scholar, either associated with the university or with the company producing them.
%NOTE: SM: this is an analysis and should not be in caption
%between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past.
%It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers. This is the opposite for the e-puck robot, distributed with an open source license.
}
\label{fig:khepuck}
\end{figure}

Along with the institutional motivation, each contributor has a personal motivation.
%Beside the institutional motivation, there is the personal motivation of each contributor.
%The personal motivation to participate to such a project is very different than the institutional one. 
When asked about their personal motivation, the people participating in the survey cited links to their professional activity and to the specific project.
%, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}). 
A few mentioned a more abstract goal like improving our society.
\iffalse
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/motivation}
\caption{Motivation of contributing to open source hardware.}
\label{fig:motivation}
\end{figure}
\fi

@@

%Another strong difference between a standard consumer product and our vision of open hardware is that Thymio should be durable.
Another strong element of our vision of open hardware is that Thymio should be durable.
%Schools do not have large budgets for technological tools and make strong investment in training of their teachers when adopting a new tool.
As schools invest in long term training of teachers, for instance, the lifetime of the products should also be as long as possible. 
The open hardware approach fits well to this requirement, as it gives to the user, or to a generic technician, better conditions to repair the system, having the specifications of all components. This is not the case of commercial robots like Edison, for instance.
Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, the battery and the main \textsc{pcb}.
Another key element in supporting repairs by end users is the documentation of calibration methods. 
When choosing very low cost components, one faces large dispersion of characteristics. 
For example, in the Thymio the right and left wheel motors can differ in their electrical characteristics, resulting in the robot not going straight for similar speed commands to both wheels.
To correct this problem, we introduced factory calibration. %, storing in the microcontroller memory correction factors that are specific for each robot. 
To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motor.
In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.

@@

%XXXXX present a better overview of the significant improvements that the design choices bring in relation to the design goals  XXXXXXX
The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
Most of the current educational robotic activities capitalize upon one strong element, for instance the technical innovation, but miss to match the formal education requirements.
%Because they do not consider all elements at stake, these tools fail to meet some challenges of formal education.
%On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
%The result is the Thymio robot and its surrounding open source community including engineers, art designers, production and sales people, and teachers.
%The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.

>>>>

\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{2,3}$% <-this % stops a space
\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools,'' and by the EU-FP7 project ASSISIbf, no. 601074 (survey on open hardware).
Many thanks to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, and Philippe R\'etornaz are with the Laboratoire de Syst\`emes Robotiques,
        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland.
        {General rule for EPFL emails: \tt\small firstname.lastname@epfl.ch}}%
\thanks{$^{2}$Michael Bonani, Fanny Riedo and St\'ephane Magnenat are with Mobsya Association, Renens, Switzerland.
        {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
\thanks{$^{3}$St\'ephane Magnenat is also with Game Technology Center, ETH Zurich, Switzerland. 
        {\tt\small stephane@magnenat.net}}%
}

@@

Many publications present educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}. 
Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
As a result, 90\% of publications about validation of educational results have been based on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ US\$400) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment. 
%NOTE: SM: commented WeDo to save space and streamline story
%At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
%The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
%In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
%However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
Among the recent new players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49), robust, and compatible with LEGO\textsuperscript{\textregistered} bricks. 
The low price has pushed drastic design choices: very few sensors, three buttons and two \textsc{led}s as user interface, and a unidirectional communication with the computer by audio jack.
These choices strongly limit its possible usage.
%Among widely used robots, none is part of an open source hardware project.

@@

Teachers, from primary school up to high school, are a primary target user group of the Thymio project.
They decide which tools are used in their class and are key people in the education ecosystem.
For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
Among them, the availability of materials and training plays a key role.
%The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
%Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
%These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
%However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
The development of educational material and courses to train teachers requires a huge effort, based on a good mix between robotics and educational skills. 
Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
%For example, in Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
%Educational material and the related training must therefore be adapted to each situation.
%Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
A crowd-sourcing approach may solve this problem; An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO\textsuperscript{\textregistered} itself is moving this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction similar to that of open hardware projects.
An open source community regrouping developers, manufacturers, and end users is therefore a very interesting model to address the distributed development and sharing of educational material and the diffusion of training sessions.
%This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
In this paper, we study a case of implementation of this model.

@@

The resulting Thymio robot possesses a large number of simple sensors:
seven horizontal proximity sensors, two infrared sensors pointing to the ground, a three-axis accelerometer, a thermistor, and a microphone.
Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
%A microphone can record sounds and detect claps. 
%Finally, a thermistor measures temperature.
Most of these devices cost less then US\$0.20, the most expensive being the accelerometer with a cost of about US\$0.80, which is an acceptable price given the possibilities it brings to the robot.
%The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force.

@@

Our choice of electronic components implies different degrees of automatization in the production. 
Full assembly is required in order for the robots to be certified for use by children. 
%A final assembly by the user is excluded by the certification required in the case of use by children.
As a full automatization requires investments that are beyond the possibilities of this project, the current production combines automatization for most components of the \textsc{pcb} and manual operations for the final assembly, and is performed in China due to low cost of manual work. 
We have thus far produced more than 16,000 robots in batches of 2,000 units, with a cost per robot of US\$39.
The strict quality control, the management of the production, the after-sales support, part of the development costs of the software, and the margins
for distributors result in a final selling price of US\$130.

%Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
%Full assembly is required in order for the robots to be certified for use by children. 
%A final assembly by the user is excluded by the certification required in the case of use by children. 
%As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
%We therefore decided to produce in China, where manual work is less expensive.
%This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
%We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
%Because manual work is expensive in western countries, we produce through subcontractors in China.
%Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
%We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
%The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
%Moreover, when selecting features we took into consideration the complexity of the injection mold.
%The design of the mechanical parts has been aligned with these choicesthat are injected, resulting in all custom parts of Thymio costing less than 4\,\$.
%In these conditions Thymio costs 35\,\$. 
The strict quality control, the production management, the support to the users, and the margins for distributors result in a final selling price of US\$130.

@@

Several design choices, such as the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne.\footnote{\url{http://www.ecal.ch}}
The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the LEGO\textsuperscript{\textregistered} robots. The look of Edison, designed after Thymio, is also technical due to its transparent cover. 
These latter two robots implicitly target a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience.
%The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.

@@

\subsection{Programming Environments}
\label{sec:aseba}

@@

\section{Open Source Hardware, Choices and Impact}

The final key design choice among the seven mentioned in the previous section is the open source project.
This choice has an impact on the robot design and the way this device is used in the community of users. 
In this section, we analyze in more detail the implications of this choice in the context of educational robotics.
We compare these implications with the results of a survey we performed among several open source hardware communities. we collected 35 answers from 11 project leaders, 13 core design team members, 8 contributors, and 3 enthusiastic users of open hardware projects.
%people active in various open source hardware projects based on a worldwide call for contributions. 
%Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
%Although 35 answer are not many, the high proportion of project leaders and core design members allows to extract some meaningful observations.
%54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
%For instance, it is interesting to observe that 34 of the 35 answers come from male respondents.

@@

%We can distinguish two levels of motivation, the institutional and the personal one.
%The institution initiating a project is mainly looking for recognition and/or money. 
Our group has good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware strategy.
Both targeted similar users and have been sold in similar quantities.
What we can observe after more than 10 years is that Khepera generated royalties for the university, but the name of the robot was mostly associated with the name of the company producing it, not the university that developed it (see Fig.~\ref{fig:khepuck}).
The e-puck robot, with open hardware and an image better linked to the university, generated no income for the university but much more relative visibility.
%Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
In the case of Thymio, the institutional motivation was toward visibility more than money. 
Therefore, from an institutional point of view, the open source hardware strategy seemed more adapted to the desired outcomes.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/others_robots}
\caption{Comparison between e-puck and Khepera in terms of mentions in Google Scholar, either associated with the university or with the company producing them.
%NOTE: SM: this is an analysis and should not be in caption
%between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past.
%It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers. This is the opposite for the e-puck robot, distributed with an open source license.
}
\label{fig:khepuck}
\end{figure}

Along with the institutional motivation, each contributor has a personal motivation.
%Beside the institutional motivation, there is the personal motivation of each contributor.
%The personal motivation to participate to such a project is very different than the institutional one. 
When asked about their personal motivation, the people participating in the survey cited links to their professional activity and to the specific project.
%, followed by a more general motivation by the nature of the project and finally by the engineering practice (Fig.~\ref{fig:motivation}). 
A few mentioned a more general goal like improving our society.
\iffalse
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/motivation}
\caption{Motivation of contributing to open source hardware.}
\label{fig:motivation}
\end{figure}
\fi

@@

%Another strong difference between a standard consumer product and our vision of open hardware is that Thymio should be durable.
Another strong element of our vision of open hardware is that Thymio should be durable.
%Schools do not have large budgets for technological tools and make strong investment in training of their teachers when adopting a new tool.
As schools invest in long term training of teachers, for instance, the lifetime of the products should also be as long as possible. 
The open hardware approach fits well to this requirement, as it gives to the user, or to a generic technician, better conditions to repair the system, having the specifications of all components. This is not the case of proprietary robots like Edison, for instance.
Supporting this type of operation has an impact on the robot design; for instance Thymio can be easily opened with four standard screws, and we introduced connectors between key elements such as motors, speaker, the battery and the main \textsc{pcb}.
Another key element in supporting repairs by end users is the documentation of calibration methods. 
When choosing very low cost components, one faces large dispersion of characteristics. 
For example, in the Thymio the right and left wheel motors can differ in their electrical characteristics, resulting in the robot not going straight for similar speed commands to both wheels.
To correct this problem, we introduced factory calibration. %, storing in the microcontroller memory correction factors that are specific for each robot. 
To allow the user to replace a broken motor, it is essential to give him also the possibility to re-calibrate the robot and adjust the parameters of the new motor.
In Thymio, this results in the design and the documentation of calibration processes that can be performed by anyone, getting close to the original definition of open source hardware.

@@

The introduction of robots in formal education is a very challenging task, not only because of technical requirements such as low cost and interactivity, but also because of factors depending on the school environment, such as the diversity of the educational programs, the dependence on local structures and languages, or the required training of teachers.
Most of the current educational robotic activities capitalize upon one strong element, for instance the technical innovation, but miss to match the formal education requirements.
%Because they do not consider all elements at stake, these tools fail to meet some challenges of formal education.
%On the contrary, in this paper we presented a global approach that tackles these questions in a holistic way, based on open source and non-profit principles.
%The result is the Thymio robot and its surrounding open source community including engineers, art designers, production and sales people, and teachers.
%The open source hardware and the related contributions are split in two categories: the robot itself, produced by a non-profit organization and requiring very advanced skills in both design and production, and the accessories for activities, based on techniques broadly accessible and with a larger panel of contributors.
